\section{Method: MERIT}

\aca{} amortizes the \rit{} estimate $\phirit$ of the same target $\phicf$ audited
in Figure~\ref{fig:diagnostic-baseline}, producing a per-event prediction $\phihat$;
it introduces no new quantity. \method{} adds three components to the standard
retrieve--execute--write loop in an architecturally linear dependency,
\rit{}~$\rightarrow$~\aca{}~$\rightarrow$~consumers
(Figure~\ref{fig:method-architecture}). All thresholds and coefficients below are
provisional hyperparameters.

\input{figures/placeholders/fig3_method_architecture}

\subsection{Randomized Interventional Trials}
\rit{} continuously produces interventional outcome-difference labels at low budget.
With sampling probability $p = \tbd{RIT_SAMPLING_PROB_PERCENT}$ (a provisional
hyperparameter; the source default is $5\%$), a replayable task enters the trial
group. For each audited memory, \rit{} performs the paired leave-one-out comparison
of Section~4 as a \emph{shadow} (non-mutating) intervention with the frozen formal
$K$ (Stage-1 pilot $K = 5$) and the neutral pad, and emits a quadruple
$(q, m_i, \text{trajectory}, \phirit_i)$ routed to one of four disjoint label roles
(fit, calibration, development, and a sealed final audit; Section~4). Under the
neutral-pad equivalence precondition the label estimates the removal effect;
otherwise it estimates the pad-replacement contribution. Sampling runs only on
tasks that satisfy AS2. Because the naive alternative---letting the agent
self-judge whether a memory helped---can reproduce a self-confirmation trap, we
retain it only as the ablation variant A-judge rather than as the label source.

\subsection{Amortized Counterfactual Attribution}
\aca{} compresses the training labels into a per-event prediction $\phihat$, scored
in $O(d)$ per retrieved memory at a fixed feature dimension $d$. The feature vector
$x_i(q)$ has three groups: a query representation, a memory representation (a frozen
embedding encoder), and \emph{usage-behavior} features that distinguish a memory
that was retrieved but ignored from one that was actually used---$n$-gram and edit
overlap between memory content and output, explicit citation markers, context
position and retrieval rank, and the log-likelihood gain on memory tokens. A
two-layer perceptron regresses $\phihat \in [-1, 1]$ under a Huber loss
($\delta_{\mathrm{H}} = 0.1$, provisional; distinct from the SR equivalence band
$\delta$ and the governance dead-zone). Because the consumers change the closed-loop
retrieval distribution, \aca{} performs \emph{periodic recalibration} every $100$
tasks (planned configuration), fine-tuning on the newest \emph{fit}-role labels and
applying isotonic calibration on the disjoint \emph{calibration}-role labels;
provisional thresholds use the \emph{development} role, and the \emph{sealed
final-audit} pool is never used for fitting, calibration, threshold selection, or
stopping. Event-level fidelity is measured by
$\mathrm{CCC}(\phihat_{i,t}, \phirit_{i,t})$ on that sealed final-audit split. At
inference, \aca{} scores each of the top-$k$ retrieved memories and updates an
exponential moving average of the running credit $\phibar_i$ and the
positive/negative query prototype sets. The trainable parameter count is
\tbd{ACA_TRAINABLE_PARAMS}. Where a backbone exposes no token log-probabilities
(the closed-source row), the log-likelihood feature is masked and the remaining
features are used.

\subsection{Two consumers of one signal}
Governance reads the running credit $\phibar$; scope gating reads $\phibar$ (in its
retrieval score) together with the positive/negative query prototypes, which are
learned from the event-level prediction $\phihat$. Both are consumers of the single
\aca{} signal, not independent methods.

\paragraph{Credit governance.}
Three threshold rules update retention, each based on an uncertainty-aware bound
rather than a bare point estimate. A memory is \emph{evicted} when a conservative
confidence bound on $\phibar_i$ (its calibrated interval, not the point value)
remains within or below the governance dead-zone across repeated updates; two
memories are \emph{merged} when their embedding similarity exceeds $0.9$ and both
have credit intervals confidently above $0$, with the merged content inherited
unchanged from the baseline writer and only the merge \emph{trigger} credit-based
(write-time content quality is out of scope); and new memories are
\emph{quarantined}, receiving a UCB-style forced exploration bonus over their first
$n_{\min} = 3$ retrievals that does not enter the
ranking score, to avoid cold-start death. The $\pm 0.02$ governance dead-zone is a
governance hyperparameter only (not the SR equivalence band $\delta$), deliberate
because the sign of the estimate is least stable near $\phicf \approx 0$; the
confidence level and the required number of repeated updates before an irreversible
eviction are locked before running (\emph{USER\_APPROVAL\_REQUIRED}), keeping
governance consistent with the CI-based audit policy. The prior minimum-evidence
guard $n_i \ge 8$ is retained. Ablation
A2 uses the identical \aca{} architecture, features, update rule and compute budget
but swaps the interventional label for a matched observational/outcome label,
isolating interventional supervision at matched capacity; it is a matched control,
not a causal-mediation proof.

\paragraph{Scope-gated retrieval.}
Each memory maintains positive and negative query prototypes (the mean embedding
of queries with $\phihat > \tau^{+}$ and $\phihat < \tau^{-}$). The retrieval score
combines semantic relevance, scope agreement, and credit,
\begin{equation}
\text{score} = \alpha\cdot\text{rel} + \beta\big(\mathrm{sim}(q,\text{proto}^{+})
  - \mathrm{sim}(q,\text{proto}^{-})\big) + \gamma\cdot\phibar_i,
\end{equation}
with $(\alpha,\beta,\gamma) = (1.0, 0.5, 0.3)$ as provisional defaults. This is
intended to address cross-task interference by learning \emph{where} each memory
applies. The same $\phihat$ can also replace a manager's outcome reward, which we
evaluate later as a signal-portability check rather than a separate contribution.

\subsection{Cost and complexity}
The dependency $\rit{} \rightarrow \aca{} \rightarrow \text{consumers}$ is an
architecturally linear chain. \aca{} inference is $O(kd)$ per query (top-$k$
memories at feature dimension $d$), i.e. $O(d)$ per retrieved-memory event and
independent of bank size; \rit{} label collection and periodic recalibration are
amortized over the stream and accounted for separately. Constant-per-event scoring
is a design property; the end-to-end token and memory overhead---%
\tbd{TOKEN_OVERHEAD_PERCENT} tokens and \tbd{ACA_VRAM_OVERHEAD_GB} of resident
VRAM---the per-event latency \tbd{ACA_SCORING_LATENCY_MS}, and the amortized \rit{}
cost \tbd{RIT_ROLLOUT_EQUIV_PER_EVENT} remain planned measurements.
Algorithm~\ref{alg:merit} summarizes the loop.

\begin{algorithm}[t]
\caption{\method{} main loop}
\label{alg:merit}
\textbf{Input}: stream $\{q_t\}$, bank $\bank$, retriever $\ret$, policy
$\policy$, attributor $\attr$, sampling budget $p$, forced-inclusion rate $\eta$\\
\textbf{State}: fit / calibration / development / sealed-audit label pools, running
credit $\phibar$, prototypes $\text{proto}^{\pm}$, counts $n$, quarantine set,
calibration state\\
\textbf{Init}: empty pools; $\attr$ warm-started on pilot labels; new memories
quarantined
\begin{algorithmic}[1]
\FOR{$t = 1, 2, \dots$}
  \STATE $C \gets \ret_t(q_t;\bank_t,\phibar,S)$ \COMMENT{scope-gated score; quarantine bonus excluded from ranking}
  \STATE with prob.\ $\eta$, add one random eligible memory to $C$ \COMMENT{forced exploration, distinct from the quarantine bonus}
  \STATE $y_t, \text{traj} \gets \policy(q_t, C)$;\quad $r_t \gets \text{env/judge}(y_t)$
  \FOR{$m_i \in C$}
    \STATE $\phihat_i \gets \attr(x_i(q_t, \text{traj}))$; update $\phibar_i, \text{proto}^{\pm}_i, n_i$
  \ENDFOR
  \STATE $\bank \gets \text{write}(\bank, q_t, \text{traj}, r_t)$ \COMMENT{baseline writer; content design out of scope}
  \IF{$\text{replayable}(q_t)$ \AND $\text{rand}() < p$}
    \FOR{$m_i$ in probability-sampled subset of $C$}
      \STATE $\phirit \gets$ shadow paired-LOO$(q_t, C, m_i, K)$ \COMMENT{non-mutating: does not alter $\bank$}
      \STATE route $(q_t, m_i, \text{traj}, \phirit)$ to a fit / calibration / development / sealed-audit pool
    \ENDFOR
  \ENDIF
  \STATE governance$(\bank)$: evict / merge / quarantine by uncertainty-aware rules on the $\phibar$ interval and $n$
  \IF{$t \bmod 100 = 0$}
    \STATE recalibrate $\attr$: fine-tune on fit pool, isotonic on calibration pool
  \ENDIF
\ENDFOR
\end{algorithmic}
\end{algorithm}
